---
title: "Data Cleaning"
format: html
---

```{r}
#| message: false  

# Libaries
library(tidyverse)

# file names
datadir_raw <- "data/raw/"

species_file <- "species.csv"

snowsurvey_file <- "ASDN_Snow_survey.csv"
```


## Data Exploration

Import the snow survey

```{r}
# Import the species
snowsurvey_csv <- read_csv(file.path(datadir_raw, snowsurvey_file))

glimpse(snowsurvey_csv)

```
Ok, the types are not what we were expecting for the percentages of cover. Let's find out where it is coming from:

```{r}
snowsurvey_csv %>% 
  count(Snow_cover)
```
Let's focus on the non-numeric values:

```{r}
snowsurvey_csv %>% 
  count(Snow_cover) %>%
  filter(is.na(as.numeric(Snow_cover)))
```

Ok, we found our problematic values that are not numeric. There are a non-negligible number of cells with a dot as value. There is no mention of using this symbol in the metadata. We should probably have a look at those rows

```{r}
snowsurvey_csv %>% 
  filter(Snow_cover == ".") %>% 
  View()
```

Interestingly, when there is a "dot" for snow cover, it is also the case for all the other covers. Let's replace them all with NA since there is no supplemental information in the provided metadata

```{r}
snowsurvey_fixed <- snowsurvey_csv %>% 
  # filter(Snow_cover == ".") %>% 
  mutate(across(ends_with("_cover"), ~ifelse(.x == ".", NA, .x)))
```

We will now tackle the other problematic values:

The problem is similarwith "-", let's set it to NA

```{r}
snowsurvey_fixed <- snowsurvey_fixed %>% 
  # filter(Snow_cover == "-") %>%
  mutate(across(ends_with("_cover"), ~ifelse(.x == "-", NA, .x)))
```

"n/a" is pretty clear regarding how to fix it:

```{r}
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Snow_cover = ifelse(Snow_cover == "n/a", NA, Snow_cover))
```

"unk" is probably an abbreviation for unknown:

```{r}
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Snow_cover = ifelse(Snow_cover == "unk", NA, Snow_cover))
```

Finally we will set "<1" as zero:

```{r}
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Snow_cover = ifelse(Snow_cover == "<1", "0", Snow_cover))
```

Now we can test if we can transform the column:

```{r}
snowsurvey_fixed %>% 
  count(Snow_cover) %>%
  filter(is.na(as.numeric(Snow_cover)))
```

Ok we can do the transformation

```{r}
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Snow_cover = as.numeric(Snow_cover))
```

Yeah we have finally a numeric column. Now we can verify that all the values are between 0 and 100:

```{r}
snowsurvey_fixed %>% 
  filter(Snow_cover > 100) 
```

We have two values above 100, with an interesting 470%! ☃️

```{r}
snowsurvey_fixed %>% 
  filter(Snow_cover < 0) 
```

No negative value detected.










Import the species csv files with the bird species information:

```{r}
# Import the species
species_csv <- read_csv(paste0(datadir_raw, species_file))

glimpse(species_csv)
```